Widespread nitrous oxide undersaturation in U.S. lakes and reservoirs

### 2023-10-18 manuscript/manuscript_file.Rmd
10/18/2023 manuscript revision

-   incorporate in Methods section how/why size is incorporated as categorical variable and continuos variable.
-   Citation "epa_data" isn't populating citation 11 in bibliography. The bibTex entry is classified as a "website". Roy, can you attempt to make this citation play nice with nature_geosciences.csl?
-   Letters vs article format at Nature Geosciences
-   Add quantitative NO3 values to Fig.3B and Fig. 5B legend.
-   nitrate in methods
-   Lester Luan and Jana Compton for internal tech reviews.

@@

Lakes, ponds, and reservoirs are estimated to be globally important sources of nitrous oxide (N~2~O) to the atmosphere [@DelSontro_etal_2018a], but the relative lack of published N~2~O emission rates and recent evidence of N~2~O uptake across a broad range of waterbodies[@Soued_etal_2016; @Webb_etal_2019; @Kortelainen_etal_2020] have called the accuracy of emission estimates into question [@Lauerwald_etal_2019]. Here we measured dissolved N~2~O concentrations and calculated emission rates for `r dg %>% filter(sitetype == "PROB") %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` waterbodies distributed across the nine major ecoregions of the conterminous U.S. (CONUS). We use a Bayesian hierarchical approach to model the data and predicted N~2~O in`r all_predictions %>% distinct(.row) %>% summarize(nlake = n()) %>% pull()` CONUS waterbodies.  We found that N~2~O undersaturation was widespread through the CONUS, with an estimated `r propN2oSinkMean` % of the waterbodies functioning as sinks for atmospheric N~2~O. Despite widespread N~2~O uptake by CONUS waterbodies, a small proportion of the waterbodies in each ecoregion had sufficiently high N~2~O emission rates to offset N~2~O uptake in the majority of waterbodies, and none of the ecoregions are estimated to functione as net N~2~O sinks. The model predicts that waterbodies with high nitrate (NO~3~) tend to be sources of N~2~O whereas those with low NO~3~ are generally sinks. Nitrate interacted with waterbody size, however, with large waterbodies having N~2~O concentrations closer to equilibrium than small waterbodies. Despite working with the largest aquatic N~2~O dataset to date, our national-scale estimate of N~2~O emissions from CONUS waterbodies is poorly constrained, with a 95% credible interval ranging from net uptake to net emission (`r national_stats %>% select(LCL_fn2o) %>% pull() %>% round(., 0)` - `r national_stats %>% select(UCL_fn2o) %>% pull() %>% round(., 0)` ) metric tons N~2~O year^-1^. The mean estimate is a net emission of  `r national_stats %>% select(post_med_fn2o) %>% pull() %>% round(., 0)` metric tons N~2~O year^-1^, which is small relative to other known sources of N2O in the CONUS (e.g., wastewater treatment, manure management) and equivalent to only `r (national_stats %>% select(UCL_fn2o) %>% pull() / 22767) * 100`% of N~2~O emissions from all CONUS surface waters (e.g. zero order streams through river mouths) per IPCC estimates . Emission estimates were most uncertain for large waterbodies, suggesting that efforts to improve the accuracy of the lentic N~2~O flux estimate should include additional measurements at the nation’s largest lakes and reservoirs.  Widespread N~2~O undersaturation in CONUS waterbodies and a national-scale emission estimate that is too uncertain to determine if CONUS waterbodies are a net source of sink of N~2~O highlight the need to revisit N~2~O models which presume surface waters are a N~2~O source [@epa_2023].

@@

While several measurement campaigns have demonstrated that waterbodies can function as N~2~O sinks[@Soued_etal_2016; @Webb_etal_2019; @Kortelainen_etal_2020; @Beaulieu_etal_2015; @Beaulieu_etal_2014; @Musenze_etal_2014], these studies have been conducted at limited spatial scales and encompassed small ranges in potential environmental drivers (e.g. nitrogen loading). Most work on aquatic N~2~O biogeochemistry has focused on streams and rivers, but these systems have been consistently found to be N~2~O sources [@Beaulieu_etal_2008; @Beaulieu_etal_2010; @Quick_etal_2019] and the controls on N~2~O biogeochmistry likely differ between flowing and standing waters. These limitations in data availability have complicated efforts to incorporate N~2~O consumption into freshwater N~2~O budgets.

@@

In this study we present the first national-scale survey of N~2~O concentrations and emissions from CONUS waterbodies. A generalized random-tessellation stratified survey design [@Stevens_Olsen_2004] was used to select a spatially-balanced and representative sample of `r dg %>% filter(sitetype == "PROB") %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` sites from CONUS waterbodies that are least 1 hectare in area, at least 1 meter deep, have at least 0.1 hectare of open water, and have a minimum residence time of one week[@EPAdesign] **[(Fig. 1)]**. Sample sites were distributed across all nine major CONUS ecoregions and varied widely in water chemistry, morphometry, watershed land-use, and climate [@epa_data]. `r undersatN2oPercent` percent of the sampled waterbodies were undersaturated in N~2~O and were therefore functioning as N~2~O sinks **[(Fig. 1)]**. We modeled the data using a Bayesian hierarchical approach and predicted the dissolved N~2~O concentration for all 465,897 CONUS waterbodies in the target population. At the population level, `r propN2oSinkMean` percent (95% CI: `r propN2oSinkLower` - `r propN2oSinkUpper`) of CONUS waterbodies were functioning as N~2~O sinks. The Western Mountains ecoregion had the smallest proportion of waterbodies functioning as N~2~O sinks (mean = `r propN2oSinkMeanWM`%, 95% CI: `r propN2oSinkLowerWM` - `r propN2oSinkUpperWM`) and the Northern Plains had the greatest (mean = `r propN2oSinkMeanNPL`%, 95% CI: `r propN2oSinkLowerNPL` - `r propN2oSinkUpperNPL`) **[(Table 1)]**.

@@

The extent of N~2~O undersaturation in CONUS waterbodies is surprisingly high, particularly for waterbodies in nitrogen rich ecoregions that are managed for agricultural production. This finding runs counter to the long standing paradigm that surface waters are a source of N~2~O to the atmosphere [@Mosier_etal_1998] and joins other recent reports that lentic waterbodies can function as an N~2~O sink. Webb et al. [@Webb_etal_2019] found that 67% of sampled reservoirs in an agricultural region of Canada were undersaturated with N~2~O during the summer months. An investigation of boreal systems in Canada found that 40% of rivers, lakes, and ponds sampled during the summer were functioning as N~2~O sinks[@Soued_etal_2016]. Similarly, 29% of surveyed boreal lakes in Finland were found to be N~2~O sinks during the summer [@Kortelainen_etal_2020]. While these latter surveys were conducted at regional scales, the national-scale results presented here provide the strongest evidence to date that most lentic waterbodies function as N~2~O sinks during the summer months. This pattern holds across all 9 major ecoregions in the CONUS, despite the tremendous variation in climate, geology, waterbody morphology, and land use across the continental U.S..

@@

Despite widespread N~2~O undersaturation in CONUS waterbodies, the mean N~2~O saturation ratio (ratio of measured to equilibrium concentration) was equal to or greater than 1 in seven of the nine ecoregions **[(Table 1, Fig. 2)]**, suggesting that, on average, waterbodies in these ecoregions were supersatured with N~2~O and were therefore functioning as a source of N~2~O. The apparent contradiction between widespread N~2~O undersaturation throughout the CONUS and average dissolved N~2~O concentrations that exceed equilibrium in most ecoregions can be attributed to the distribution of dissolved N~2~O concentrations across the population of CONUS waterbodies **[(Fig. 2)]**. Across the CONUS, and within each of the nine major ecoregions, the distribution of N~2~O saturation ratios had median values that were less than 1, indicating undersaturation, but a long right tail composed of relatively rare waterbodies with levels of N~2~O supersaturation that are sufficiently high to drive the population mean to values close to, or even greater than 1. Thus most waterbodies in the CONUS are functioning as N~2~O sinks during the summer months, but a few function as strong N~2~O sources, particularly at high NO~3~ at water temperature **[Fig. 3]**.

![Figure 2. Distribution of N~2~O saturation values for the population of waterbodies in the nine major ecoregions. Each blue line (n=2000) represents the distribution of values from a single realization of the posterior prediction. Vertical black dashed and solid lines indicate median and mean, respectively. Vertical red line indicates a saturation ratio of 1. x-axis is truncated at 3 to better show the area where 97% of the observations occur.](manuscript_figures/figure2.tiff)

@@

It is also possible that the correlation between N~2~O and NO~3~ reflects the simultaneous production of these solutes via nitrification. Nitrification, an aerobic process which can produce N~2~O as a byproduct, is the microbial oxidation of ammonium (NH~4~^+^) to NO~3~ [@Kuypers_etal_2018]. Nitrification might be expected to be more prevalent in the predominantly well oxygenated surface waters where the N~2~O samples were collected than denitrification, which is an anaeorbic process. However, denitrification can occur in anoxic microsites within particulate matter suspended in an aerobic water column [@Broman_etal_2021] and can be transported from aoxic sediments to the water surface via advection and diffusion [@DelSontro_etal_2018]. Thus the N~2~O - NO~3~ relationship may be reflective of some combination of N~2~O production/consumption via denitrification and N~2~O production via nitrification.

![Figure 3. Conditional effects plot illustrating A) the relationship between N~2~O saturation and nitrate (NO~3~) concentration, dependent upon surface water temperature. The five NO~3~ categories correspond to concentration ($\mu$g L^-1^) ranges of 1) below detection, 2) $\geq$ 0.55 and \< 4.09, 3) $\geq$ 4.09 and \< 30.20, 4) $\geq$ 30.20 and \< 223.13, and 5) $\geq$ 223.13, and B) relationship between N~2~O saturation and surface water temperature, dependent upon NO~3~ category. Horizontal dashed lines in both panels reflects an N~2~O saturation value of 1.](manuscript_figures/sat_cond_surftemp.tiff)

@@

N~2~O saturation also increased with water temperature **[Fig. 3]**, which is consistent with N~2~O production and consumption occurring through temperature sensitive biological processes, but the relationship was dependent on NO~3~ availability. At low NO~3~ concentrations, temperature had a relatively minor affect on N~2~O saturation, possibly because N~2~O production/consumption was limited by substrate rather than temperature. The effect of temperature became more pronounced when NO~3~ availability was greater, likely reflecting the stimulatory affect of higher temperature on biological N~2~O production. The relative effect of NO~3~ was greater than that of temperature, thus elevated water temperature serves to magnify rates of N~2~O production that are largely dictated by NO~3~ availability.

@@

While NO~3~ and water temperature were the primary drivers of N~2~O production or consumption rates, dissolved N~2~O concentration was also related to waterbody size. When pooled across all ecoregions, the absolute magnitude of the difference between the measured and equilibrium N~2~O concentration, known as $\Delta$ N~2~O, decreases with increasing waterbody size class **[Fig. 4A]**. This pattern holds regardless of whether N~2~O is supersaturated or undersatured **[Fig. 4B, C]**. Waterbody size and shape can affect dissolved gas concentration through several mechanisms. In lakes with low water volume to benthic surface area ratios, the mass of N~2~O consumed or produced in the sediments can be a substantial fraction of the mass of N~2~O dissolved in the water column [@Carpenter_1983]. Furthermore, sediment may warm faster in small and shallow lakes, whereas thermal stratification may substantially delay sediment warming in large and deep lakes, resulting in greater rates of N~2~O production or consumption in small lakes. Thermal stratification may also prevent the exchange of water between the lake surface and the hypolimnion, thereby preventing N~2~O depleted or enriched water from advecting to the lake surface [@Beaulieu_etal_2015; @Beaulieu_etal_2014].

Persistent patterns in the air-water gas exchange rate across lake size classes may also contribute to greater N~2~O disequilibrium in smaller waterbodies. Air-water gas exchange will vent excess N~2~O from surface waters or facilitate the invasion of air that can replenish depleted N~2~O in undersaturated waterbodies. Regardless of whether a waterbody is superaturated or undersatured in N~2~O, air-water gas exchange will function to move the dissolved N~2~O concentration toward equilibrium. One of the primary drivers of air-water gas exchange in lentic waters is wind speed [@Vachon_etal_2013]. Wind speeds tend to be greater in large lakes with long fetches and lower in small waterbodies, particularly if they are sheltered by shoreline vegetation. Thus, greater air-water gas exchange rates in large lakes may maintain dissolved N~2~O concentrations close to equilibrium, whereas the affect of biogeochemistry on dissolved N~2~O is more pronounced in small waterbodies because low air-water gas exchange limits the rate at which excess N~2~O can be ventilated or depleted N~2~O can be replenished. While previous surveys have documented relationships between lake size and dissolved methane concentration [@Pighini_etal_2018; @Juutinen_etal_2009; @Borges_etal_2011], to the best of our knowledge this is the first study to document a relationship between lake size and $\Delta$ N~2~O.

![Figure 4. A) Relationship between $\Delta$ N~2~O (difference between observed and equilibrium N~2~O concentration) and waterbody surface areas. B,C) relationship between N~2~O saturation and waterbody size for the Coastal Plains and Northern Plains ecoregions. The vertical blue line references a N~2~O saturation value of 1.0. The horizontal red arrows highlight $\Delta$ N~2~O for the smallest lake size class. Mean (points) and 95% credible interview (horizontal bars) are derived from 2000 realizations of the poterior predictive distribution.](manuscript_figures/deltaN2ObySize.tiff)

The influence of waterbody size is also evident in the statistical interaction between NO~3~ and surface area **[Fig. 5]**. At high NO~3~ concentrations, N~2~O is highly supersatured in small waterbodies, but moves toward equilibrium in large waterbodies **[Fig. 5A]**. At low NO~3~ concentrations, N~2~O is undersatured in small waterbodies, but increases toward equilibrium in larger waterbodies **[Fig. 5A]**.

@@

The model indicates that NO~3~ exerts a strong control on N~2~O, but the relationship is complicated by interactions with water temperature and lake surface area. This finding mirrors reports from a survey of farm ponds in Canada where N~2~O was found to be partially controlled by dissolved inorganic nitrogen (DIN) concentration, but the effect of DIN was conditional upon and thermal stratification strength. These complex interactions may explain contrasting literature reports regarding the role of NO~3~ in regulating N~2~O. Several studies have reported positive correlations between NO~3~ and N~2~O in lakes [@Kortelainen_etal_2020; @Yang_etal_2015] while others have found no relationship [@Soued_etal_2016; @Whitfield_etal_2011]. It is possible that the NO~3~ \~ N~2~O relationship wasn't detected in the latter studies because it was confounded by interactions with temperature, lake size, thermal stratification, or other environmental variables.

Interactions between NO~3~, water temperature, and lake size produced a broad range of N~2~O saturation values across the population of CONUS waterbodies, but also enable the identifcation of broad scale patterns. First, small waterbodies with low NO~3~ concentration tend to be undersaturated, whereas small waterbodies with high NO~3~ tend to be highly supersaturated **[Fig. 5]**. Large waterbodies, regardless of NO~3~ status, tend to have N~2~O saturation ratios closer to equilibrium. Therefore intra and inter regional patterns in N~2~O saturation can be well predicted by the joint distributions of waterbody size, NO~3~ availability, and water temperature.

@@

The emission rates reported here span a broad range (`r min.max$min %>% round(.,2)` - `r min.max$max %>% round(.,2)` mg N~2~O m^-2^ d^-1^), which is expected for this large and diverse collection of waterbodies, but are within the range of values reported in the literature for lentic ecosystems [@Yang_etal_2015; @Kortelainen_etal_2020; @Webb_etal_2019]. Variability and magnitude of predicted N~2~O emission rates decreased with increasing waterbody size **[(Fig. 6A)]**, which is consistent with the relationship between N~2~O saturation and waterbody size **[(Fig. 5A,B)]**. Despite widespread N~2~O undersaturation in CONUS waterbodies, the national average N~2~O emission rate was positive (mean = `r national_stats$post_med_en2o`, 95% CI: `r national_stats$LCL_en2o` - `r national_stats$UCL_en2o` mg N~2~O m^-2^ d^-1^) **[(Table 1)]**. This contradiction is due to relatively rare waterbodies with very high emission rates that pull the mean (`r national_stats$post_med_en2o` mg N~2~O m^-2^ d^-1^) to a value that far exceeds the median (`r national_stats$post_median_en2o` mg N~2~O m^-2^ d^-1^) **[(SI Fig. 1)]**.

![Figure 6: A) N~2~O emission rate and B) flux plotted against waterbody size for all 465,896 waterbodies in the target population. C) N~2~O flux plotted against the five lake size categories used in the survey design. Across all plots, points and bars reflect the mean (points) and 95% credible interval of the 2000 realizations of the posterior predictive distribution.](manuscript_figures/n2oFluxAndEmissionRateVsContinuousArea.tiff)

@@

Our survey is the first to demonstrate that most lakes and reservoirs in the CONUS function as N~2~O sinks during the summer. Despite widespread N~2~O uptake, CONUS waterbodies function as a net source of N~2~O due to high emissions from from a subset of the population. Dissolved N~2~O in lakes and reservoirs can be predicted from non-linear interactions between water temperature, NO~3~ availability, and waterbody size. NO~3~ availability largely determines whether a waterbody functions as a sink or source of N~2~O, and the magnitude of the source or sink is exaggerated with increasing water temperature. The degree of N~2~O disequilibrium also increases with decreasing waterbody size, possibly because smaller waterbodies have low air-water gas exchange rates.

>>>>

-   incorporate in Methods section how/why size is incorporated as categorical variable and continuos variable.
-   Citation "epa_data" isn't populating citation 11 in bibliography. The bibTex entry is classified as a "website". Roy, can you attempt to make this citation play nice with nature_geosciences.csl?
-   Letters vs article format at Nature Geosciences
    + Letters no longer supported.  Of all the options, Article is best suited to us.
    + 2200 – 3000 words (excluding abstract, Methods, references and figure legends)
    + Pseudo-abstract, 200 words, no citations
    + Subheadings are OK
    + Instructions to authors list separate Results and Discussion, but article in current issue do not follow this format.
    + Figures: article in current issue have 2-5 figures, mostly multi-panel.  Instructions specify 4-6 figures + tables.
-   Add quantitative NO3 values to Fig.3B and Fig. 5B legend.
-   Lester Luan and Jana Compton for internal tech reviews.

@@

**Lakes, ponds, and reservoirs are estimated to be globally important sources of nitrous oxide (N~2~O) to the atmosphere [@DelSontro_etal_2018a], but the relative lack of published N~2~O emission rates and recent evidence of N~2~O uptake across a broad range of waterbodies[@Soued_etal_2016; @Webb_etal_2019; @Kortelainen_etal_2020] have called the accuracy of emission estimates into question [@Lauerwald_etal_2019]. Here we measured N~2~O concentrations and calculated emission rates for `r dg %>% filter(sitetype == "PROB") %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` waterbodies distributed across the nine major ecoregions of the conterminous U.S. (CONUS). We use a Bayesian hierarchical approach to model the data and predicted N~2~O in`r all_predictions %>% distinct(.row) %>% summarize(nlake = n()) %>% pull()` CONUS waterbodies.  We found that N~2~O undersaturation was widespread through the CONUS, with an estimated `r propN2oSinkMean` % (95% credible interval:`r propN2oSinkLower` - `r propN2oSinkUpper`% ) of waterbodies functioning as sinks for atmospheric N~2~O. Despite widespread N~2~O uptake by CONUS waterbodies, a small proportion of the waterbodies in each ecoregion had sufficiently high N~2~O emission rates to offset N~2~O uptake in the majority of waterbodies, and none of the ecoregions are estimated to function as net N~2~O sinks. The model predicts that waterbodies with high nitrate (NO~3~) tend to be sources of N~2~O whereas those with low NO~3~ are generally sinks. Nitrate interacted with waterbody size, however, with large waterbodies having N~2~O concentrations closer to equilibrium than small waterbodies. Despite working with the largest aquatic N~2~O dataset to date, our national-scale estimate of N~2~O emissions from CONUS waterbodies is poorly constrained, with a 95% credible interval ranging from net uptake to net emission (`r national_stats %>% select(LCL_fn2o) %>% pull() %>% round(., 0)` - `r national_stats %>% select(UCL_fn2o) %>% pull() %>% round(., 0)` ) metric tons N~2~O year^-1^. The mean estimate is a net emission of `r national_stats %>% select(post_med_fn2o) %>% pull() %>% round(., 0)` metric tons N~2~O year^-1^, which is small relative to other known sources of N2O in the CONUS (e.g., wastewater treatment, manure management) and equivalent to only `r (national_stats %>% select(UCL_fn2o) %>% pull() / 22767) * 100 %>% round(., 1)`% of N~2~O emissions from all CONUS surface waters (e.g. zero order streams through river mouths) per IPCC estimates . Emission estimates were most uncertain for large waterbodies, suggesting that efforts to improve the accuracy of the lentic N~2~O flux estimate should include additional measurements at the nation’s largest lakes and reservoirs.  Widespread N~2~O undersaturation in CONUS waterbodies and a national-scale emission estimate that is too uncertain to determine if CONUS waterbodies are a net source or sink of N~2~O highlight the need to revisit N~2~O models which presume surface waters are a N~2~O source.**

@@

While several measurement campaigns have demonstrated that waterbodies can function as N~2~O sinks[@Soued_etal_2016; @Webb_etal_2019; @Kortelainen_etal_2020; @Beaulieu_etal_2015; @Beaulieu_etal_2014; @Musenze_etal_2014], these studies have been conducted at limited spatial scales and encompassed small ranges in potential environmental drivers (e.g. nitrogen loading). Nitrous oxide dynamics have been investigated in streams and rivers spanning a broad range of environemtal conditions, but these systems have been consistently found to be N~2~O sources [@Beaulieu_etal_2008; @Beaulieu_etal_2010; @Quick_etal_2019] and the controls on N~2~O biogeochmistry likely differ between flowing and standing waters. These limitations in data availability have complicated efforts to incorporate N~2~O consumption into freshwater N~2~O budgets.

@@

In this study we present the first national-scale survey of N~2~O concentrations and emissions from CONUS waterbodies. A generalized random-tessellation stratified survey design [@Stevens_Olsen_2004] was used to select a spatially-balanced and representative sample of `r dg %>% filter(sitetype == "PROB") %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` sites from CONUS waterbodies that were at least 1 hectare in area, at least 1 meter deep, had at least 0.1 hectare of open water, and had a minimum residence time of one week[@EPAdesign] **[(Fig. 1)]**. Sample sites were distributed across all nine major CONUS ecoregions and varied widely in water chemistry, morphometry, watershed land-use, and climate [@epa_data]. `r undersatN2oPercent` percent of the sampled waterbodies were undersaturated in N~2~O and were therefore functioning as N~2~O sinks **[(Fig. 1)]**. These data, along with the survey design information, were used to develop a Bayesian hierarchical model to predict N~2~O concentrations for all 465,897 CONUS waterbodies in the target population. At the population level, the majority of CONUS waterbodies (mean = `r propN2oSinkMean`%, 95% CI: `r propN2oSinkLower` - `r propN2oSinkUpper`) were predicted to be a net sink for N~2~O. The Western Mountains ecoregion had the smallest proportion of waterbodies functioning as N~2~O sinks (mean = `r propN2oSinkMeanWM`%, 95% CI: `r propN2oSinkLowerWM` - `r propN2oSinkUpperWM`) and the Northern Plains had the greatest (mean = `r propN2oSinkMeanNPL`%, 95% CI: `r propN2oSinkLowerNPL` - `r propN2oSinkUpperNPL`), although the 95% credible intervals overlap across all ecoregions **[(Table 1)]**.

@@

The estimated extent of N~2~O undersaturation in CONUS waterbodies was surprisingly high, particularly for waterbodies in nitrogen rich ecoregions that are managed for agricultural production. This finding runs counter to the long-standing paradigm that surface waters are a source of N~2~O to the atmosphere [@Mosier_etal_1998] and joins other recent reports that lentic waterbodies can function as N~2~O sinks. Webb et al. [@Webb_etal_2019] found that 67% of sampled reservoirs in an agricultural region of Canada were undersaturated with N~2~O during the summer months. An investigation of boreal systems in Canada found that 40% of rivers, lakes, and ponds sampled during the summer were functioning as N~2~O sinks[@Soued_etal_2016]. Similarly, 29% of surveyed boreal lakes in Finland were found to be N~2~O sinks during the summer [@Kortelainen_etal_2020]. While these latter surveys were conducted at regional scales, the national-scale results presented here provide the strongest evidence to date that most lentic waterbodies function as N~2~O sinks during the summer months. This pattern held across all 9 major ecoregions in the CONUS, despite the tremendous variation in climate, geology, waterbody morphology, and land use.

@@

Despite widespread N~2~O undersaturation in CONUS waterbodies, the mean predicted N~2~O saturation ratio (ratio of measured to equilibrium concentration) was statistically indistinguishable from one in all ecoregions **[(Table 1, Fig. 2)]**. The apparent contradiction between widespread N~2~O undersaturation throughout the CONUS and average dissolved N~2~O concentrations close to equilibrium in all ecoregions can be attributed to the predicted distribution of dissolved N~2~O concentrations across the population of CONUS waterbodies **[(Fig. 2)]**. Across the CONUS, and within each of the nine major ecoregions, the predicted distribution of N~2~O saturation ratios had median values less than 1, indicating undersaturation, but a long right tail composed of relatively rare waterbodies with levels of N~2~O supersaturation that are sufficiently high to drive the population mean to values close to, or even greater than 1. Thus most waterbodies in the CONUS are functioning as N~2~O sinks during the summer months, but a few function as strong N~2~O sources, particularly at high NO~3~ at high water temperature **[Fig. 3]**.

![Figure 2. Posterior predictive distribution of N~2~O saturation values for waterbodies in the nine major ecoregions. Each blue line (n=2000) represents the distribution of values from a single realization of the posterior predictive distribution. Vertical black dashed and solid lines indicate point estimates of the median and mean, respectively (Table 1). Vertical red line indicates a saturation ratio of 1. x-axis is truncated at 3 to better show the area where 97% of the observations occur.](manuscript_figures/figure2.tiff)

@@

It is also possible that the correlation between N~2~O and NO~3~ reflects the simultaneous production of these solutes via nitrification. Nitrification, an aerobic process which can produce N~2~O as a byproduct, is the microbial oxidation of ammonium (NH~4~^+^) to NO~3~ [@Kuypers_etal_2018]. Nitrification may be more prevalent in the predominantly well oxygenated surface waters where the N~2~O samples were collected because denitrification is an anaeorbic process. However, denitrification can occur in anoxic microsites within particulate matter suspended in an aerobic water column [@Broman_etal_2021] and can be transported from aoxic sediments to the water surface via advection and diffusion [@DelSontro_etal_2018]. Thus the N~2~O - NO~3~ relationship may be reflective of some combination of N~2~O production/consumption via denitrification and N~2~O production via nitrification.

![Figure 3. Predicted interactive effects of surface water temperature and NO~3~ category the N~2~O saturation ratio.  A) The relationship between N~2~O saturation and nitrate (NO~3~) concentration, dependent upon surface water temperature. The five NO~3~ categories correspond to concentration ($\mu$g L^-1^) ranges of 1) below detection, 2) $\geq$ 0.55 and \< 4.09, 3) $\geq$ 4.09 and \< 30.20, 4) $\geq$ 30.20 and \< 223.13, and 5) $\geq$ 223.13, and B) relationship between N~2~O saturation and surface water temperature, dependent upon NO~3~ category. Horizontal dashed lines in both panels reflects an N~2~O saturation value of 1.](manuscript_figures/sat_cond_surftemp.tiff)

@@

The model detected a pattern of increasing N~2~O saturation with water temperature **[Fig. 3]**, which is consistent with nitrogen transformations occurring through temperature sensitive biological processes, but the relationship was dependent on NO~3~ concentration. At low NO~3~ concentrations, temperature was estimated to have a relatively minor effect on N~2~O saturation, possibly because N~2~O production/consumption was limited by substrate rather than temperature. The effect of temperature was more pronounced at greater NO~3~ concentrations, potentially reflecting the stimulatory effect of higher temperature on biological N~2~O production. The relative effect of NO~3~ was estimated to be greater than that of temperature, thus elevated water temperature may serve to magnify rates of N~2~O production that are largely dictated by NO~3~ availability.

@@

While NO~3~ and water temperature were associated with the N~2~O saturation ratio, N~2~O was also related to waterbody size. When pooled across all ecoregions, the absolute magnitude of the difference between the measured and equilibrium N~2~O concentration, known as $\Delta$ N~2~O, decreased with increasing waterbody size class **[Fig. 4]**. This pattern holds regardless of whether N~2~O was supersaturated or undersatured. Waterbody size and shape can affect dissolved gas concentration through several mechanisms. In lakes with low water volume to benthic surface area ratios, the mass of N~2~O consumed or produced in the sediments can be a substantial fraction of the mass of N~2~O dissolved in the water column [@Carpenter_1983]. Furthermore, sediment may warm faster in small and shallow lakes, whereas thermal stratification may substantially delay sediment warming in large and deep lakes, resulting in greater rates of N~2~O production or consumption in small lakes. Thermal stratification may also prevent the exchange of water between the lake surface and the hypolimnion, thereby preventing N~2~O depleted or enriched water from advecting to the lake surface [@Beaulieu_etal_2015; @Beaulieu_etal_2014].

Persistent patterns in the air-water gas exchange rate across lake size classes may also contribute to greater N~2~O disequilibrium in smaller waterbodies. Air-water gas exchange will vent excess N~2~O from surface waters or facilitate the invasion of air that can replenish depleted N~2~O in undersaturated waterbodies. Regardless of whether a waterbody is supersaturated or undersaturated in N~2~O, air-water gas exchange will function to move the dissolved N~2~O concentration toward equilibrium. One of the primary drivers of air-water gas exchange in lentic waters is wind speed [@Vachon_etal_2013]. Wind speeds tend to be greater in large lakes with long fetches and lower in small waterbodies, particularly if they are sheltered by shoreline vegetation. Thus, greater air-water gas exchange rates in large lakes may maintain dissolved N~2~O concentrations close to equilibrium, whereas the effect of biogeochemistry on dissolved N~2~O may be more pronounced in small waterbodies because low air-water gas exchange limits the rate at which excess N~2~O can be ventilated or depleted N~2~O can be replenished. While previous surveys have documented relationships between lake size and dissolved methane concentration [@Pighini_etal_2018; @Juutinen_etal_2009; @Borges_etal_2011], to the best of our knowledge this is the first study to document a relationship between lake size and $\Delta$ N~2~O.

![Figure 4. Relationship between $\Delta$ N~2~O (difference between predicted dissolved and equilibrium N~2~O concentration) and waterbody surface areas. Mean (points) and 95% credible interview (horizontal bars) are derived from 2000 realizations of the posterior predictive distribution.](manuscript_figures/deltaN2ObySize.tiff)

The influence of waterbody size is also suggested in the interaction between NO~3~ and surface area **[Fig. 5]**. At high NO~3~ concentrations, N~2~O is highly supersatured in small waterbodies, but moves toward equilibrium in large waterbodies **[Fig. 5A]**. At low NO~3~ concentrations, N~2~O is undersaturated in small waterbodies, but increases toward equilibrium in larger waterbodies **[Fig. 5A]**.

@@

The model results suggest that NO~3~ may exert a strong control on N~2~O, but the relationship is complicated by interactions with water temperature and lake surface area. This finding mirrors reports from a survey of farm ponds in Canada where N~2~O was found to be associated with dissolved inorganic nitrogen (DIN) concentration, but the effect of DIN was conditional upon thermal stratification strength. These complex interactions may explain contrasting literature reports regarding the role of NO~3~ in regulating N~2~O. Several studies have reported positive correlations between NO~3~ and N~2~O in lakes [@Kortelainen_etal_2020; @Yang_etal_2015] while others have found no relationship [@Soued_etal_2016; @Whitfield_etal_2011]. It is possible that the NO~3~ \~ N~2~O relationship wasn't detected in the latter studies because it was confounded by interactions with temperature, lake size, thermal stratification, or other environmental variables.

Interactions between NO~3~, water temperature, and lake size produced a broad range of N~2~O saturation values across the population of CONUS waterbodies, but also enabled identifcation of broad scale patterns. First, small waterbodies with low NO~3~ concentration tended to be undersaturated, whereas small waterbodies with high NO~3~ tended to be highly supersaturated **[Fig. 5]**. Large waterbodies, regardless of NO~3~ status, tended to have N~2~O saturation ratios closer to equilibrium. Therefore intra and inter regional patterns in N~2~O saturation may be well predicted by the joint distributions of waterbody size, NO~3~ availability, and water temperature.

@@

The emission rates reported here span a broad range (`r min.max$min %>% round(.,2)` - `r min.max$max %>% round(.,2)` mg N~2~O m^-2^ d^-1^), which was expected for this large and diverse collection of waterbodies, but are within the range of values reported in the literature for lentic ecosystems [@Yang_etal_2015; @Kortelainen_etal_2020; @Webb_etal_2019]. Variability and magnitude of predicted N~2~O emission rates decreased with increasing waterbody size **[(Fig. 6A)]**, which is consistent with the relationship between N~2~O saturation and waterbody size **[(Fig. 5A,B)]**. Despite predicted widespread N~2~O undersaturation in CONUS waterbodies, the estimated national average N~2~O emission rate was positive (mean = `r national_stats$post_med_en2o`, 95% CI: `r national_stats$LCL_en2o` - `r national_stats$UCL_en2o` mg N~2~O m^-2^ d^-1^) **[(Table 1)]**. This contradiction is due to relatively rare waterbodies with very high emission rates that pull the mean (`r national_stats$post_med_en2o` mg N~2~O m^-2^ d^-1^) to a value that far exceeded the median (`r national_stats$post_median_en2o` mg N~2~O m^-2^ d^-1^) **[(SI Fig. 1)]**.

![Figure 6: A) N~2~O emission rate and B) flux plotted against waterbody size for all 465,896 waterbodies in the target population. C) N~2~O flux plotted against the five lake size categories used in the survey design. Across all plots, points and bars reflect the mean (points) and 95% credible interval, respectively, of the 2000 realizations of the posterior predictive distribution.](manuscript_figures/n2oFluxAndEmissionRateVsContinuousArea.tiff)

@@

Our research is the first to suggest that most lakes and reservoirs in the CONUS function as N~2~O sinks during the summer.  Our observations of dissolved N~2~O were predicted reasonably well using non-linear interactions between observed water temperature, NO~3~ concentration, and waterbody size (see SI). The model form and estimated parameters suggest that NO~3~ concentration plays a role in determining whether a waterbody functions as a sink or source of N~2~O, and the magnitude of the source or sink is exaggerated with increasing water temperature. The degree of N~2~O disequilibrium also increases with decreasing waterbody size, possibly because smaller waterbodies have low air-water gas exchange rates. Our observations of dissolved N~2~O was predicted  due to high emissions from from a subset of the population.

Although waterbodies in the largest lake size class tend to have N~2~O levels close to equilibrium, they emit more N~2~O than the other size classes due to their expansive surface area. The N~2~O flux from these waterbodies is also more uncertain than the smaller lake size classes because relatively small uncertainties in dissolved and equilibrium N~2~O concentrations translate to large uncertainties in flux when propagated across the surface of large waterbodies. Uncertainty in the national flux estimate was too large to determine if CONUS waterbodies functioned as a net source or sink of N~2~O. While our sample included some of the largest waterbodies in the CONUS, efforts to improve the accuracy of the lentic N~2~O flux estimate should include additional measurements at the nations largest lakes including an assesment of intra-waterbody spatial variation in N~2~O, which can be important [@Salk_etal_2019, @Borges_etal_2023], as well as careful accounting of measurement an analytical uncertainty which can propagate to large uncertainties in flux estimates.
